Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-12-22 bmc_template/figures.Rnw
need turning angle picture

<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",fig.width=6,fig.height=4,eval=FALSE>>=
ICdat <- readRDS("../sim_results/ICdat.rds")
aa <- readRDS("../sim_results/sim.101.RDS")

ICdat <- mutate(ICdat,
                model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
    geom_line() +
    geom_point(size=3,bg="white") +
    scale_shape_manual(values=c(21,16))+
    scale_colour_brewer(palette="Set1")+
    labs(y="Counts",x="BIC-optimal number of states")
# print(simplot)

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4,eval=FALSE>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"Time-heterogeneous","Homogeneous"))
slname <- "Step-length\ndistribution"
(ggplot(df2, aes(x=nstates,y=deltaBIC,size=distrib,
                 colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name=slname,palette="Set1")
    + scale_size_manual(name=slname,values=c(1.75,0.75))
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, temporally heterogeneous FMM with a sinusoidal prior and HMM. The right panel shows HMMs with different temporal transitions.",fig.width=8,eval=FALSE>>=
#L <- load("cat1sumdat.RData")

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.1.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)

# print(adj_BIC_plot)
@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8,eval=FALSE>>=

avgdf <- readRDS("../summary_stats/cat.1.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.1.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.1.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("FMM-homogeneous  (n=5)"
             , "FMM-sin (n=4)"
             , "HMM-homogeneous (n=6)"
             , "HMM-block (n=4)"
             , "HMM-hourly (n=3)"
             , "HMM-quadratic (n=5)"
             , "HMM-sin (n=5)"
             , "observed"
             , "viterbi homogenous (n=3)")

combdf <- (rbind(avgdf,obsdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

@@

(avg_plot <- ggplot(combdf,aes(x=time,y=steplen,colour=model,
                   lwd=model,
                   alpha=model,linetype=model, shape=model))
  + geom_point()
  + geom_line()
  + xlab("Time of Day") 
  + scale_shape_manual(values=shapevec, name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
 )

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ",eval=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

@@

(acf_plot <- ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
                  lwd=model,alpha=model,linetype=model,shape=model))
  + geom_point()
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_shape_manual(values=shapevec, name="Model")
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
) 
# print(acf_plot)
## vitacf messes up legend ...
@

<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE,eval=FALSE>>=
get_cat_pars2 <- function(catid){
  templist <- readRDS(paste("../summary_stats/cat.",catid,".df.RDS",sep=""))
  moddf <- templist$msdlist

@@

@@

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.2.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

@@

avgdf <- readRDS("../summary_stats/cat.2.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.2.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.2.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm4","fmmsin4","hmm5","hmmblock4","hmmhourly3",
            "hmmquad4","hmmsin4","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=4)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=5)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=4)"
             , "sin (n=4)"
             , "observed"
             , "viterbi homogenous (n=3)")

combdf <- (rbind(avgdf,obsdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

@@

(ggplot(combdf,aes(x=time,y=steplen,colour=model,
                   lwd=model,
                   alpha=model,linetype=model, shape=model))
  + geom_point()
  + geom_line()
  + xlab("Time of Day") 
  + scale_shape_manual(values=shapevec, name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
 ) 
@

@@

@@

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.14.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

@@

avgdf <- readRDS("../summary_stats/cat.14.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.14.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.14.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm4","fmmsin3","hmm5","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin4","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=4)"
             , "fmm-sin (n=3)"
             , "homogeneous (n=5)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=5)"
             , "sin (n=4)"
             , "observed"
             , "viterbi homogenous (n=3)")

combdf <- (rbind(avgdf,obsdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2))
)
linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

@@

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.15.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

@@

avgdf <- readRDS("../summary_stats/cat.15.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.15.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean))
  %>% transmute(time = Time, variable = "obs", value=LogDist)
)
vit <- readRDS("../summary_stats/cat.15.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm3","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=3)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=5)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=5)"
             , "sin (n=5)"
             , "observed"
             , "viterbi homogenous (n=3)")

>>>>

<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",eval=TRUE>>=
ICdat <- readRDS("../simsin_results/ICdat.rds")

ICdat <- (ICdat 
          %>% ungroup() 
          %>% mutate(model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous")))
          %>% filter(IC == "BIC")
)
simsin <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
    geom_line() +
    geom_point(size=3,bg="white") +
    scale_shape_manual(values=c(21,16))+
    scale_colour_brewer(palette="Set1")+
    ggtitle("2-state HMM with Time-dependent Transition")+
    labs(y="Counts",x="BIC-optimal number of states")+
    theme(legend.position="none"))

ICdat <- readRDS("../sim_results/ICdat.rds")

ICdat <- (ICdat 
          %>% ungroup() 
          %>% mutate(model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous")))
          %>% filter(IC == "BIC")
)
simgps <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
    geom_line() +
    geom_point(size=3,bg="white") +
    scale_shape_manual(values=c(21,16))+
    scale_colour_brewer(palette="Set1")+
    ggtitle("2-state HMM with GPS Error")+
    labs(y="Counts",x="BIC-optimal number of states")+
      theme(legend.position="none"))

ICdat <- readRDS("../simtime_results/ICdat.rds")

ICdat <- (ICdat 
          %>% ungroup() 
          %>% mutate(model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous")))
          %>% filter(IC == "BIC")
)
simtimegps <- (ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
    geom_line() +
    geom_point(size=3,bg="white") +
    scale_shape_manual(values=c(21,16))+
    scale_colour_brewer(palette="Set1")+
    ggtitle("2-state HMM with Time-dependent Transition and GPS Error")+
    labs(y="Counts",x="BIC-optimal number of states")+
      theme(legend.position="none"))

print(grid.arrange(simsin,simgps,simtimegps,ncol=2))

@@

\clearpage
<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4,eval=TRUE>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"Time-heterogeneous","Homogeneous"))
slname <- "Step-length\ndistribution"
(ggplot(df2, aes(x=nstates,y=deltaBIC,size=distrib,
                 colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name=slname,palette="Set1")
    + scale_size_manual(name=slname,values=c(1.75,0.75))
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, temporally heterogeneous FMM with a sinusoidal prior and HMM. The right panel shows HMMs with different temporal transitions.",fig.width=8,eval=TRUE>>=
#L <- load("cat1sumdat.RData")

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.1.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   # + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point()
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   # + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)

#print(adj_BIC_plot)
@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8,eval=TRUE>>=

avgdf <- readRDS("../summary_stats/cat.1.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.1.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean,sd))
  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
)
vit <- readRDS("../summary_stats/cat.1.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("FMM-homogeneous  (n=5)"
             , "FMM-sin (n=4)"
             , "HMM-homogeneous (n=6)"
             , "HMM-block (n=4)"
             , "HMM-hourly (n=3)"
             , "HMM-quadratic (n=5)"
             , "HMM-sin (n=5)"
             , "Observed"
             , "viterbi homogenous (n=3)"
             )

combdf <- (rbind(avgdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
                      , smin=steplen,smax=steplen)
)

combdf2 <- rbind(combdf,obsdf)
linetypev <- c(rep(c(2,1),c(2,5))
                ,1,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"))
             ,"black","black"
            )

lwdvec <- c(rep(c(1),c(7))
             ,1,4
            )
alphavec <- c(rep(0,7)
               ,0.1,0
              )

@@

avg_plot <- ggplot(combdf2,aes(x=time))
print(avg_plot 
  + geom_ribbon(aes(ymin=smin,ymax=smax,alpha=model))
  + geom_line(aes(y=steplen,linetype=model,colour=model))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + geom_point(aes(y=steplen,colour=model,shape=model,lwd=model))
  + scale_shape_manual(values=shapevec,name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + xlab("Time of Day") 
)

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ",eval=TRUE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

@@

linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

(acf_plot <- ggplot(combacfdf,aes(x=lag,y=ACF,colour=model,
                  lwd=model,alpha=model,linetype=model,shape=model))
  + geom_point()
  + geom_line()
  + xlab("Lag") 
  + ylab("Autocorrelation")
  + scale_shape_manual(values=shapevec, name="Model")
  + scale_colour_manual(values=colvec, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
) 
print(acf_plot)
## vitacf messes up legend ...
@

\clearpage
<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE,eval=TRUE>>=
get_cat_pars2 <- function(catid){
  templist <- readRDS(paste("../summary_stats/cat.",catid,".df.RDS",sep=""))
  moddf <- templist$msdlist

@@

Everything Below goes into the appendix

@@

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.2.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   # + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

@@

avgdf <- readRDS("../summary_stats/cat.2.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.2.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean,sd))
  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
)

vit <- readRDS("../summary_stats/cat.2.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm4","fmmsin4","hmm5","hmmblock4","hmmhourly3",
            "hmmquad4","hmmsin4","obs","viterbi")
mnames2 <- c("Fmm-homogeneous  (n=4)"
             , "Fmm-sin (n=4)"
             , "HMM-homogeneous (n=5)"
             , "HMM-block (n=4)"
             , "HMM-hourly (n=3)"
             , "HMM-quadratic (n=4)"
             , "HMM-sin (n=4)"
             , "Observed"
             , "Viterbi homogenous (n=3)")
combdf <- (rbind(avgdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
                      , smin=steplen,smax=steplen)
)

combdf2 <- rbind(combdf,obsdf)
linetypev <- c(rep(c(2,1),c(2,5))
                ,1,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"))
             ,"black","black"
            )

lwdvec <- c(rep(c(1),c(7))
             ,1,4
            )
alphavec <- c(rep(0,7)
               ,0.1,0
              )

@@

avg_plot <- ggplot(combdf2,aes(x=time))
(avg_plot 
  + geom_ribbon(aes(ymin=smin,ymax=smax,alpha=model))
  + geom_line(aes(y=steplen,linetype=model,colour=model))
  + scale_colour_manual(values=colvec, name="Model")
  + scale_size_manual(values=lwdvec, name="Model")
  + scale_linetype_manual(values=linetypev, name="Model")
  + scale_alpha_manual(values=alphavec, name="Model")
  + geom_point(aes(y=steplen,colour=model,shape=model,lwd=model))
  + scale_shape_manual(values=shapevec,name="Model")
  + ylab(expression('Mean step length'~(log[10]*m)))
  + xlab("Time of Day") 
)

@@

linetypev <- c(rep(c(2,1),c(2,6))
                ,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"),"black")
             ,"black"
            )
lwdvec <- c(rep(c(1,4),c(7,1))
             ,4
            )
alphavec <- c(rep(c(1,0.2),c(7,1))
               ,1
              )

@@

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.14.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   # + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

@@

avgdf <- readRDS("../summary_stats/cat.14.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.14.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean,sd))
  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
)
vit <- readRDS("../summary_stats/cat.14.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm4","fmmsin3","hmm5","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin4","obs","viterbi")
mnames2 <- c("Fmm-homogeneous  (n=4)"
             , "Fmm-sin (n=3)"
             , "HMM-homogeneous (n=5)"
             , "HMM-block (n=4)"
             , "HMM-hourly (n=3)"
             , "HMM-quadratic (n=5)"
             , "HMM-sin (n=4)"
             , "Observed"
             , "Viterbi homogenous (n=3)")
combdf <- (rbind(avgdf,vitdf2)
           %>% transmute(time=time,model=variable,steplen=value)
           %>% filter(model %in% mnames)
           %>% mutate(model=factor(model,levels=mnames,labels=mnames2)
                      , smin=steplen,smax=steplen)
)

combdf2 <- rbind(combdf,obsdf)
linetypev <- c(rep(c(2,1),c(2,5))
                ,1,0
               )
colvec <- c(c(brewer.pal(7,"Dark2"))
             ,"black","black"
            )

lwdvec <- c(rep(c(1),c(7))
             ,1,4
            )
alphavec <- c(rep(0,7)
               ,0.1,0
              )

@@

linetypev <- c(rep(2,2),rep(1,5))
sumdat <- transform(readRDS("../summary_stats/cat.15.df.RDS")$HMMsumdf,
                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                  , y=deltaBIC
                                  , colour=model
                                  , linetype=model
                                  , group=model))
   + facet_wrap(~mod,ncol = 2,scales = "free_x")
   + scale_size_continuous(name="# of states")
   # + scale_x_continuous(breaks=seq(0,300,by=30))
   + geom_point(aes(shape=factor(nstates),alpha=nstates))
   + scale_linetype_manual(values=linetypev, name="Model")
   + scale_colour_brewer(palette="Dark2", name="Model")
   + scale_shape_discrete(name="Number of states")
   + scale_alpha_continuous(name="Number of states")
   + labs(y=expression(paste(Delta, "BIC"))
          , x="Number of parameters") 
   + geom_line()
   + zmargin)
@

@@

avgdf <- readRDS("../summary_stats/cat.15.df.RDS")$avgplotdf
obsdf <- (readRDS("../summary_stats/cat.15.RDS") 
  %>% dplyr::select(c(LogDist,Time)) 
  %>% filter(!is.na(LogDist)) 
  %>% group_by(Time) 
  %>% summarise_each(funs(mean,sd))
  %>% transmute(time=Time,model="Observed",steplen=mean,smin=mean-2*sd,smax=mean+2*sd)
)

vit <- readRDS("../summary_stats/cat.15.vit.RDS")
vitdf <- (vit %>% dplyr::select(c(viterbi,time))
  %>% group_by(time)
  %>% summarise_each(funs(mean))
  %>% mutate(model="homogeneous (n=3)")
)

@@

mnames <- c("fmm3","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("Fmm-homogeneous  (n=3)"
             , "Fmm-sin (n=4)"
             , "HMM-homogeneous (n=5)"
             , "HMM-block (n=4)"
             , "HMM-hourly (n=3)"
             , "HMM-quadratic (n=5)"
             , "HMM-sin (n=5)"
             , "Observed"
             , "Viterbi homogenous (n=3)")
